---
title: "Dealing with Endogeneity: Instrumental Variable"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 4:3
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, child = './../setup.Rmd'}
```

```{r echo = F, eval = F}
setwd("/Users/tmieno2/Dropbox/TeachingUNL/AppliedEconometrics_MS/LectureNotes/Endogeneity/")
```

```{r additional-libraries, include = F, cache = F}
#--- load packages ---#
suppressMessages(library(broom))
suppressMessages(library(fixest))
suppressMessages(library(stargazer))
suppressMessages(library(lmtest))
suppressMessages(library(readstata13))
```

# Before we start 

## Learning objectives

Understand how instrumental variable (IV) estimation works.

## Table of contents

1. [Instrumental Variable (IV) Approach](#inst)
2. [IV in R](#iv-r)


---
class: middle

# Endogeneity  

.content-box-red[**Endogeneity**]
  
$E[u|x_k] \ne 0$ (the error term is not correlated with any of the independent variables)

--

.content-box-red[**Endogenous independent variable**]

If the error term is, .red[for whatever reason], correlated with the independent variable $x_k$, then we say that $x_k$ is an endogenous independent variable.

+ Omitted variable 
+ Selection 
+ Reverse causality 
+ Measurement error

---
class: inverse, center, middle
name: inst

# Instrumental Variable (IV) Approach

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---
class: middle

.content-box-green[**The Model**]

$y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + u$

+ $x_1$ is endogenous: $E[u|x_1] \ne 0$ (or $Cov(u,x_1)\ne 0$)
+ $x_2$ is exogenous: $E[u|x_1] = 0$ (or $Cov(u,x_1) = 0$)

---
class: middle

.content-box-green[**Idea (very loosely put)**]

Bring in variable(s) (.blue[Instrumental variable(s)]) that does .red[NOT] belong to the model, but .red[IS] related with the endogenous variable,

+ using the instrumental variable(s) (which we denote by $Z$), make the endogenous variable exogenous, which we call .blue[instrumented variable(s)]

+ use the variation in the instrumented variable instead of the original endogenous variable to estimate the impact of the original variable

---
class: middle

# IV estimation procedure

.content-box-green[**Step 1**]
  
using the instrumental variables, make the endogenous variable exogenous, which we call .blue[instrumented variable]

--

.content-box-green[**Step 1: mathematically**]

+ Regress the endogenous variable $(x_1)$ on the instrumental variable(s) $(Z=\{z_1,z_2\}$, two instruments here) and all the other exogenous variables that you do not intend to instrument,

$x_1 = \alpha_0 + \sigma_2 x_2 + \alpha_1 z_1 +\alpha_2 z_2 + \mu$

--

+ obtain the predicted value of $x$ from the regression

$\hat{x}_1 = \hat{\alpha}_0 + \hat{\sigma}_2 x_2 + \hat{\alpha}_1 z_1 + \hat{\alpha}_2 z_2$

---
class: middle

# IV estimation procedure

.content-box-green[**Step 2**]

use the variation in the instrumented variable instead of the original endogenous variable to estimate the impact of the original variable


.content-box-green[**Step 2: Mathematically**]

Regress the dependent variable $(y)$ on the instrumented variable $(\hat{x}-1)$,

$y= \beta_0 + \beta_1 \hat{x_1}+ \beta_2 x_2 + \varepsilon$

to estimate the coefficient on $x$ in the original model

---
class: middle

# Does IV really work? 

.content-box-green[**Estimation Procedure**]

+ Step 1: $\hat{x}_1 = \hat{\alpha}_0 +\hat{\sigma}_2 x_2 + \hat{\alpha}_1 z_1 + \hat{\alpha}_2 z_2$  

+ Step 2: $y = \beta_0 + \beta_1 \hat{x_1}+ \beta_2 x_2 + \varepsilon$

--

.content-box-green[**Important question**]

What are the conditions under which IV estimation is consistent?

--

The instruments $(Z)$ need to satisfy two conditions, which we will discuss.

---
class: middle

# Condition 1

---
class: middle

.content-box-green[**Question**]

What happens if $Z$ have no power to explain $x_1$ $(E[\hat{\alpha}_1]=0$ and $E[\hat{\alpha}_2]=0)$?

--

.content-box-green[**Answer**]

+ $\hat{x}_1=\hat{\alpha}_0+\hat{\sigma}^2 x_2$
+ $\hat{\beta}_1?$

--

That is, $\hat{x_1}$ has no information beyond the information $x_2$ possesses.  

---
class: middle

.content-box-red[**Condition 1**]

The instrument(s) $Z$ have jointly significant explanatory power on the endogenous variable $x_1$ .red[after] you control for all the other exogenous variables (here $x_2)$}

---
class: middle center

# Condition 2

---
class: middle

+ Instrumentation phase: $\hat{x}_1 = \hat{\alpha}_0 +\hat{\sigma}_2 x_2 + \hat{\alpha}_1 z_1 + \hat{\alpha}_2 z_2$  

+ Main estimation phase: $y = \beta_0 + \beta_1 \hat{x_1}+ \beta_2 x_2 + \varepsilon$

--

Plugging in $x_1 = \hat{\alpha}_0 + \hat{\sigma}_2 x_2 + \hat{\alpha}_1 z_1 +\hat{\alpha}_2 z_2 + \hat{\mu}$ into the original model,

$y =  \beta_0 + \beta_1 (\hat{\alpha}_0 + \hat{\sigma}_2 x_2 + \hat{\alpha}_1 z_1 +\hat{\alpha}_2 z_2 + \hat{\mu}) + u$ 
$\;\;\;=  \beta_0 + \beta_1 \times \underline{(\hat{\alpha}_0 + \hat{\sigma}_2 x_2 + \hat{\alpha}_1 z_1 +\hat{\alpha}_2 z_2)} +\beta_2 x_2 + v \;\; (=\beta_1 \hat{\mu} + u)$

--

.content-box-green[**Question**]

What is the condition under which the OLS estimation of $\beta_1$ in the main model is unbiased?

--

.content-box-green[**Answer**]

$(\hat{\alpha}_0 + \hat{\sigma}_2 x_2 + \hat{\alpha}_1 z_1 +\hat{\alpha}_2 z_2)$ is not correlated with $v = \beta_1\mu + u$

--

.content-box-red[**Condition 2**]

$z_1$ and $z_2$ are not correlated with $\mu$ and $u$.

---
class: middle

.content-box-red[**Condition 2**]

$z_1$ and $z_2$ are not correlated with $\mu$ and $u$.

--

.content-box-green[**Question**]

When is this satisfied?

---
class: middle

.content-box-green[**Question**]

Are $z_1$ and $z_2$ uncorrelated with $\mu$? 

--

.content-box-green[**Answer**]

Always yes. Independent variables and the residuals are orthogonal (uncorrelated).

--

.content-box-green[**Question**]

When are $z_1$ and $z_2$ uncorrelated with $u$? 

--

.content-box-green[**Answer**]

+ $z_1$ and $z_2$ do not belong in the main model, meaning they do not have any explanatory power beyond $x_2$ (they should have been included in the model in the first place as independent variables)

+ $z_1$ and $z_2$ are not correlated with the error term (there are no unobserved factors in the error term that are correlated with $Z)$

---
class: middle

.content-box-green[**Question**]

Do you think we can test condition 2?

--

.content-box-green[**Answer**]

No, because we never observe the error term.

--

.content-box-red[**Important**]

+ All we can do is to .red[argue] that the instruments are not correlated with the error term.

--

+ In journal articles that use IV method, they make careful arguments as to why their choice of instruments are not correlated with the error term.


---
class: middle

.content-box-red[**Condition 1**]

+ The instrument(s) $Z$ have jointly significant explanatory power on the endogenous variable $x_1$ .red[after] you control for all the other exogenous variables (here $x_2)$}


.content-box-red[**Condition 2**]

+ $z_1$ and $z_2$ do not belong in the main model, meaning they do not have any explanatory power beyond $x_2$ (they should have been included in the model in the first place as independent variables)

+ $z_1$ and $z_2$ are not correlated with the error term (there are no unobserved factors in the error term that are correlated with $Z)$

--

.content-box-red[**Important**]

+ Condition 1 is always testable

+ Condition 2 is NOT testable (unless you have more instruments than endogenous variables)

---
class: middle


.content-box-green[**Two-stage Least Square (2SLS)**]

IV estimator is also called two-stage least squares estimator (2SLS) because it involves two stages of OLS.

+ Step 1: $\hat{x}_1 = \hat{\alpha}_0 +\hat{\sigma}_2 x_2 + \hat{\alpha}_1 z_1 + \hat{\alpha}_2 z_2$  

+ Step 2: $y = \beta_0 + \beta_1 \hat{x_1}+ \beta_2 x_2 + \varepsilon$     

--

+ 2SLS framework is a good way to understand conceptually why and how instrumental variable estimation works

+ But, IV estimation is done in one-step

---
class: middle

# Instrumental variable validity

---
class: middle

.content-box-green[**The model**]

$log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + v \;\; ( = \beta_3 ability + u)$

`educ` is endogenous because of its correlation with `ability`.

--

.content-box-green[**Question**]

What conditions would a good instrument $(z)$ satisfy?

--

.content-box-green[**Answer**]

+ $z$ has explanatory power on $educ$ \textcolor{blue}{after} you control for the impact of $epxer$ on $educ$

+ $z$ is uncorrelated with $v$ ($ability$ and all the other important unobservables)

---
class: middle

.content-box-green[**The model**]

$log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + v \;\; ( = \beta_3 ability + u)$

.content-box-green[**An example of instruments**]

The last digit of an individual's Social Security Number? (this has been actually used in some journal articles)

.content-box-green[**Question**]

+ Is it uncorrelated with $v$ ($ability$ and all the other important unobservables)?

--

+ does it have explanatory power on $educ$ .blue[after] you control for the impact of $epxer$ on $educ$?

---
class: middle

.content-box-green[**The model**]

$log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + v \;\; ( = \beta_3 ability + u)$

.content-box-green[**An example of instruments**]

IQ test score

.content-box-green[**Question**]

+ Is it uncorrelated with $v$ ($ability$ and all the other important unobservables)?

--

+ does it have explanatory power on $educ$ .blue[after] you control for the impact of $epxer$ on $educ$?

---
class: middle

.content-box-green[**The model**]

$log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + v \;\; ( = \beta_3 ability + u)$

.content-box-green[**An example of instruments**]

Mother's education

.content-box-green[**Question**]

+ Is it uncorrelated with $v$ ($ability$ and all the other important unobservables)?

--

+ does it have explanatory power on $educ$ .blue[after] you control for the impact of $epxer$ on $educ$?

---
class: middle

.content-box-green[**The model**]

$log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + v \;\; ( = \beta_3 ability + u)$

.content-box-green[**An example of instruments**]

Number of siblings

.content-box-green[**Question**]

+ Is it uncorrelated with $v$ ($ability$ and all the other important unobservables)?

--

+ does it have explanatory power on $educ$ .blue[after] you control for the impact of $epxer$ on $educ$?


---
class: middle

class: inverse, center, middle
name: iv-r

# Implementation of Instrumental Variable (IV) Estimation in R

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---
class: middle

.content-box-green[**Model**]

$log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + v \;\; (=\beta_3 ability + u)$

We believe

+ $educ$ is endogenous $(x_1)$

+ $exper$ is exogenous $(x_2)$

+ we use the number of siblings ($sibs$) and father's education ($feduc$) as the instruments ($Z$)

--

.content-box-green[**Terminology**]

+ exogenous variable included in the model (here, $exper$) is also called .blue[included instruments]

+ instruments that do not belong to the main model (here, $sibs$ and $feduc$) are also called .blue[excluded instruments]

+ we refer to the collection of included and excluded instruments as .blue[instruments]

---
class: middle


.content-box-green[**Dataset**]

```{r }
#--- read the data ---#   
data <- read.dta13('WAGE2.dta')

#--- take a look at the data ---#
select(data, wage, educ, sibs, feduc) %>% head
```

---
class: middle

We can use the `lfe` package to run IV estimation method.

```{r cache = F}
library(lfe)  
```

---
class: middle 

.content-box-green[**Syntax**]

```{r eval = F}
lfe(dep var ~ included instruments|0|first stage formula|0 , data = dataset)  
```

+ `included instruments`: exogenous included variables (do not include endogenous variables here)

--

.content-box-green[**first stage formula**]

```{r eval = F}
(endogenous vars ~ excluded instruments)
```

--

.content-box-green[**Example**]

```{r }
iv_res <- felm(log(wage) ~ exper| 0 | (educ ~ sibs + feduc) | 0, data = data)  
```

+ `included variable`: 
  * exogenous included variables: `exper`
  * endogenous included variables: `educ` 
+ `instruments`: 
  * included instruments: `exper` 
  * excluded instruments: `sibs` and `feduc`

---
class: middle

```{r }
tidy(iv_res)  
```

---
class: middle


.content-box-green[**Comparison of OLS and IV Estimation Results**]

```{r echo = F, results = "asis"}
#--- OLS ---#  
reg_ols <- lm(log(wage)~educ+exper,data=data)

#--- Stargaze ---#  
stargazer(reg_ols,iv_res,single.row=TRUE,omit.stat=c('all'),table.layout='-ldc-t-',type='html',column.labels=c('OLS','IV'))  
```

--


.content-box-green[**Question**]

Do you think $sibs$ and $feduc$ are good instruments?

+ Condition 1: weak instruments?
+ Condition 2: uncorrelated with the error term?

---
class: middle

.content-box-green[**Weak Instrument Test**]

We can always test if the excluded instruments are weak or not (test of condition 1).

--

.content-box-green[**How**]

+ Run the 1st stage regression

$educ = \alpha_0 + \alpha_1 exper + \alpha_2 sibs + \alpha_3 feduc + v$  

--

+ test the joint significance of $\alpha_2$ and $\alpha_3$ ($F$-test)

If excluded instruments $(sibs$ and $feduc$) are jointly significant, then it would mean that $sibs$ and $feduc$ are not weak instruments, satisfying condition 1.

---
class: middle

When we ran the IV estimation using `iv_reg()` earlier, it automatically calculated the F-statistic for the weak instrument test.

To see it, access the `stage1` element of `iv_res`, and further access the `rob.iv1fstat` component of `stage1`.

--

```{r }
iv_res$stage1$rob.iv1fstat
```

+ `F`: F-statistic for the weak instrument test
+ `P`: p-value for the weak instrument test

---
class: middle

.content-box-green[**Notes**]

+ It is generally recommended that you have $F$-stat of over $10$ (this is not a clear-cut criteria that applied to all the empirical cases)

+ Even if you reject the null if $F$-stat is small, you may have a problem

+ You know nothing about if your excluded instruments satisfy Condition II. 

+ If you cannot reject the null, it is a strong indication that your instruments are weak. Look for other instruments.

+ Always, always report this test. There is no reason not to. 

---
class: middle

# Consequences of weak instruments 

.content-box-green[**Data generation**]

```{r }
set.seed(73289)
N <- 500 # number of observations  

u_common <- runif(N)# the term shared by the endogenous variable and the error term 
z_common <- runif(N) # the term shared by the endogenous variable and instruments
x_end <- u_common + z_common + runif(N) # the endogenous variable 
z_strong <- z_common + runif(N) # strong instrument
z_weak <- 0.01*z_common + 0.99995*runif(N) # weak instrument
u <- u_common + runif(N) # error term
y <- x_end + u # dependent variable

data <- data.frame(y,x_end,z_strong,z_weak)
```

---
class: middle

.content-box-green[**Correlation**]

```{r }
cor(data)
```

---
class: middle


.content-box-green[**Estimation with the strong instrumental variable**]

```{r }
#--- IV estimation (strong) ---#
iv_strong <- felm(y ~ 1 | 0 | (x_end ~ z_strong) | 0, data = data)
```

.content-box-green[**Estimation with the weak instrumental variable**]

```{r }
#--- IV estimation (weak) ---#
iv_weak <- felm(y ~ 1 | 0 | (x_end ~ z_weak) | 0, data = data)
```

---
class: middle

```{r }
#--- coefs (strong) ---#
tidy(iv_strong)

#--- coefs (weak) ---#
tidy(iv_weak)  
```


.content-box-green[**Question**]

Any notable differences?

--

The coefficient estimate on $x\_end$ is far away from the true value in the weak instrument case.

---
class: middle


.content-box-green[**Comparison of the weak instrument tests**]

.scroll-box-10[
```{r }
#--- diagnostics (strong) ---#
iv_strong$stage1$rob.iv1fstat
```
]

.scroll-box-10[
```{r }
#--- diagnostics (weak) ---#
iv_weak$stage1$rob.iv1fstat 
```
]

.content-box-green[**Question**]

Any notable differences?

--

You cannot reject the null hypothesis of weak instrument in the weak instrument case.

---
class: middle


.content-box-green[**MC simulation**]

```{r }
B <- 1000 # the number of experiments
beta_hat_store <- matrix(0,B,2) # storage of beta hat

for (i in 1:B){

  #--- data generation ---#
  u_common <- runif(N) 
  z_common <- runif(N) 
  x_end <- u_common + z_common + runif(N) 
  z_strong <- z_common + runif(N)
  z_weak <- 0.01*z_common + 0.99995*runif(N)
  u <- u_common + runif(N)
  y <- x_end + u
  data <- data.table(y,x_end,z_strong,z_weak)

  #--- IV estimation with a strong instrument ---#
  iv_strong <- felm(y ~ 1 | 0 | (x_end ~ z_strong) | 0, data=data)
  beta_hat_store[i,1] <- iv_strong$coef[2, 1]

  #--- IV estimation with a weak instrument ---#
  iv_weak <- felm(y ~ 1 | 0 | (x_end ~ z_weak) | 0, data=data)
  beta_hat_store[i,2] <- iv_weak$coef[2, 1]
}  
```

---
class: middle

.content-box-green[**Visualization of the MC Results**]

```{r echo = F, out.width = "70%"}
melted <- melt(data.table(beta_hat_store))
melted[variable=='V1',variable:='Strong']
melted[variable=='V2',variable:='Weak']

ggplot(data=melted) +
  geom_density(aes(x=value,fill=variable),alpha=0.3) +
  geom_vline(xintercept=1,color='red') +
  scale_fill_discrete(name='') +
  theme(
  legend.position='bottom'
  )   
```

---
class: middle

.content-box-green[**Visualization of the MC Results**]

```{r echo = F, out.width = "70%"}
ggplot(data=melted[abs(value)<3,]) +
  geom_density(aes(x=value,fill=variable),alpha=0.3) +
  geom_vline(xintercept=1,color='red') +
  scale_fill_discrete(name='') +
  theme(legend.position='bottom')   
```

---
class: middle

.content-box-green[**Flow of IV Estimation in Practice**]

+ Identify endogenous variable(s) and included instrument(s) 

+ Identify potential excluded instrument(s) 

+ .red[Argue] why the excluded instrument(s) you pick is uncorrelated with the error term (.content-box-red[**condition 2**])

+ Once you decide what variable(s) to use as excluded instruments, .red[test] whether the excluded instrument(s) is weak or not (
.content-box-red[**condition 1**])

+ Implement IV estimation and report the results
