---
title: "Research Demonstration: The Impacts of Irrigation Costs on Irrigation Demand"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    self_contained: true
    nature:
      ratio: 4:3
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r, child = './../setup.Rmd'}
```

```{r echo = F, eval = F}
setwd(here("LectureNotes/14_ResearchDemonstration"))
```

```{r additional-libraries, include = F, cache = F}
#--- load packages ---#
library(broom)
library(fixest)
library(flextable)
library(officer)
library(gt)
library(lmtest)
library(lfe)
library(tidyverse)
library(data.table)
```

.content-box-green[**Objective**]

Understand the impact of pumping cost on groundwater use for irrigated agriculture

<br>

.content-box-green[**Data**]

Well-level groundwater use observed annually in Eastern Colorado

---
class: middle

# Data

```{r read-data}
(data <- readRDS("final_data_CO.rds") %>%
  dplyr::select(site_no, year, pumpingAF, pc, precip, et, gpm, REA, tier))
```

+ `site_no`: well id 
+ `year`: year 
+ `pumpingAF`: groundwater use (acre-feet)
+ `pc`: pumping cost ($/acre-feet)
+ `precip`: total precipitation (inches) during the growing season
+ `et`: total evapotranspiration (inches) during the growing season
+ `gpm`: well yield (gallons per minute)
+ `REA`: energy supplier name
+ `tier`: price tier (explained later)

---
class: middle

# Exploratory Data Analysis

---

class: middle

# Econometric Model (univariate)

$W = \beta_0 + \beta_1 IC + v$

+ $W$: irrigation amount 
+ $IC$: irrigation cost 
+ $v$: error term.

---
class: middle

# Endogeneity: Omitted Variable Bias

$W_{i,t} = \beta_0 + \beta_1 IC_{i,t} + v_{i,t}$

<br>

.content-box-green[**Question**]

What is in $v$? Are they going to cause bias on $\beta_1$ estimation? 

--

+ soil type
+ well yield
+ precipitation 

---
class: middle

# Econometric Model (multi-variate) 

$W_{i,t} = \beta_0 + \beta_1 IC_{i,t} + \beta_2 WY_{i,t} + \beta_3 Precip_{i,t} + \beta_4 ET_{i,t} + \varepsilon_{i,t}$

+ $W$: irrigation amount 
+ $IC$: irrigation cost 
+ $WY$: well yield
+ $Precip$: total precipitation during the production season
+ $ET$: total evapotranspiration during the production season
+ $\varepsilon$: error term

---
class: middle

# Can Individual (well) Fixed Effects help?

---
class: middle

# Endogeneity

.content-box-red[**Declining block rate pricing**]

```{r echo = F}
block_price_table <-
  fread("REA_price_schedule.csv") %>%
  .[REA == "Highline" & year >= 2013, t2 := NA] %>%
  .[REA == "Highline" & year >= 2013, p3 := NA] %>%
  .[year <= 2016, ] %>%
  .[, REAnum := NULL] %>%
  .[, REA := NULL] %>%
  setnames(names(.), c("Year", "Price 1", "Price 2", "Price 3", "Threshold 1", "Threshold 2")) %>%
  .[, Supplier := "Highline"] %>%
  .[7:12, Supplier := "Y-W"] %>%
  relocate(Supplier) %>%
  flextable() %>%
  merge_v(j = 1) %>%
  hline(
    i = 6,
    border = fp_border(width = 1)
  ) %>%
  fix_border_issues() %>%
  autofit()
```

```{r echo = F}
block_price_table
```

Threshold 1 and 2 refer to the threshold electricity uses (kwh/HP) over which the users move on to the next tier. For example, for a farmer served by Highline who has a pump of 100 HP, the farmer would be at the second tier after using 40,000 kwh.

---
class: middle

# Endogeneity

.content-box-red[**Declining block rate pricing**]

```{r echo = F}
block_price_table
```

.content-box-green[**Example**]

A well in Highline's service area with 100 HP will be on the second price tier after using 40,000 kWh.

---
class: middle

# Endogeneity

.content-box-red[**Declining block rate pricing**]

```{r echo = F}
block_price_table
```

.content-box-green[**Endogeneity?**]

Would this cause bias in estimating $\beta_1$?

---
class: middle

# Fixed-Effects estimation 

```{r echo = F}
massage_data_2p <- data.frame(
  Location = rep(c("Chicago", "Peoria", "Milwaukee", "Madison"), each = 2),
  Year = rep(c(2003, 2004), 4),
  P = c(75, 85, 50, 48, 60, 65, 55, 60),
  Q = c(2, 1.8, 1.0, 1.1, 1.5, 1.4, 0.8, 0.7),
  Ql = rep(c(10, 5, 7, 6), each = 2)
)
```

```{r echo = F}
massage_data_2p %>%
  flextable()
```

---
class: middle

# (individual) Fixed Effects Estimation

+ Including individual (cross-sectional units) dummies (here, they are wells represented by `wdid`) in the model

+ This is equivalent to do within-transformation of the data (deviation from the individual means) and then regress deviation on deviation

+ What the within-transformation does is to effectively  divide the observations into groups (wells) and you look at variations .red[within] each of the groups (wells observed over time).

+ Since including individual dummies are equivalent to the within-transformation approach, that means including individual (well) dummies also effectively divides the observations into groups (wells) and you look at variations .red[within] each of the groups (wells observed over time).

---
class: middle

# Fixed Effect Estimation (in general)

.content-box-green[**key**]

Including dummies variables effectively divide the observations into groups and you look at variations .red[within] each of the groups

---
class: middle

# Beyond individual (well) FEs  

```{r echo = F}

final_data[wdid %in% c(4905049, 4905044), .(wdid, year, pumpingAF, pc, tier, pump_hp)] %>%
  arrange(wdid, year) %>%
  flextable() %>%
  merge_v(j = 1) %>%
  hline(
    i = 5,
    border = fp_border(width = 1)
  ) %>%
  fix_border_issues() %>%
  autofit()
```

--

.content-box-green[**Question**]

+ Does including individual dummies (well fixed effects) help?

--

+ Does including well-tier dummies (well-tier fixed effects) help? 

--

+ Does including well-tier-hp dummies (well-tier-hp fixed effects) help? 
