---
title: "Univariate Regression: Part 1"
author: "AECN 896-003"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Table of contents

1. [What is Rmarkdown?](#intro)
2. [Chunk options](#objects)
3. [Functions and packages](#functions)
4. [Some fundamentals on vector, matrix, list, and data.frame](#objects_basics)

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  header_h1_font_size = "2.1rem",
  header_h2_font_size = "1.5rem",
  code_inline_background_color = "#e6e6e6", 
  text_font_size = "0.9rem",
  code_font_size = "0.8rem",
  link_color = "#ff005a"
)

```

```{r , include=FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center", #fig.width=6, fig.height=4.5, 
  # out.width="748px", #out.length="520.75px",
  # dpi = 100, #fig.path='Figs/',
  cache = T#, echo=F, warning=F, message=F
  )

opts_knit$set(root.dir = "~/Box/Teaching/UNL/AppliedEconometrics_MS/LectureNotes")
# setwd("~/Box/Teaching/UNL/DataScience") 
```

```{r, include = F}
#--- load packages ---#
suppressMessages(library(data.table))
suppressMessages(library(exactextractr))
suppressMessages(library(prism))
suppressMessages(library(sf))
suppressMessages(library(raster))
suppressMessages(library(tidyverse))
suppressMessages(library(DT))
suppressMessages(library(tictoc))
suppressMessages(library(tmap))
suppressMessages(library(stargazer))
suppressMessages(library(xtable))
suppressMessages(library(parallel))
suppressMessages(library(maps))
````


class: inverse, center, middle
name: intro

# Univariate Regression

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<!-- 
#/*=================================================*/
#' # Basic Data Transformation
#/*=================================================*/
-->

---

# Plan

+ simple univariate regression analysis for the next two weeks
+ multivariate regression after that

---

# Population and Sample 

## Population

A set of $ALL$ individuals, items, phenomenon, that you are interested in learning about

--

## Example
+ Suppose you are interested in the impact of eduction on income across the U.S. Then, the population is all the individuals in U.S.

+ Suppose you are interested in the impact of water pricing on irrigation water demand for farmers in NE. Then, your population is all the farmers in NE.

--

## Important

Population differs depending on the scope of your interest

---


# Population and Sample

## Sample

Sample is a subset of population that you observe

--

## Example  
+ data on education, income, and many other things for 300 individuals from each State

+ data on water price, irrigation water use, and many other things for 500 farmers who farm in the Upper Republican Basin (southwest corner of NE)

--

## Question

Are the samples representative of the population you are interested in learning about?

---

# Econometrics 

Econometrics is about learning the population using sample

---

# Simple linear regression model

Consider a phenomenon in the population that is correctly represented by the following model <span style="color:blue"> (This is the model you want to learn about using sample) </span>

$$y=\beta_0+\beta_1 x + u$$

+ $y$: to be explained by $x$ (\textcolor{blue}{dependent variable})

+ $x$: explain $y$ (\textcolor{blue}{independent variable}, \textcolor{blue}{covariate})

+ $u$: parts of $y$ that cannot be explained by $x$ (\textcolor{blue}{error term})

+ $\beta_0$ and $\beta_1$: real numbers that gives the model a quantitative meaning (\textcolor{blue}{parameters})



