---
title: "Summary Questions"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    self_contained: true
    nature:
      ratio: 4:3
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r, child = './../setup.Rmd'}
```

```{r echo = F, eval = F}
setwd("/Users/tmieno2/Dropbox/TeachingUNL/AppliedEconometrics_MS/LectureNotes/Endogeneity/")
```

```{r additional-libraries, include = F, cache = F}
#--- load packages ---#
library(broom)
library(fixest)
library(lmtest)
```

# Interpretation of an econometric model

Consider the following model:

$$
\begin{aligned}
y_i = \alpha + \beta_1 x_{1,i} + \beta_2 x^2_{1,i} + \beta_3 x_{1,i} \times x_{2,i} + \beta_4 log(x_{3,i}) +v_{i}
\end{aligned}
$$

+ $y$ is the dependent variable
+ $x_{1}$, $x_{2}$, and $x_{3}$ are explanatory variables 
+ $v$ is the error term

<br>

.content-box-green[**Questions**]

+ What is the marginal impact of $x_i$?
+ How do you interpret $\beta_4$?
+ What does $\alpha$ represent? 

---
class: middle

# Small and large sample properties

.content-box-green[**Questions**]

+ What is the difference between estimators and estimates?
+ What does "unbiasedness" of an estimator mean?
+ What does "consistency" of an estimator mean?
+ Write the mathematical expression of the variance of the OLS estimator of the coefficient on $x_j$, where $x_j$ is one of the explanatory variables in a multivariate regression model.
+ Inspect the variance of the OLS estimator you wrote above and discuss what affects it.

---
class: middle

# Unbiasedness of OLS estimator of the coefficients

.content-box-green[**Questions**]

+ What are the four conditions that you need to satisfy for the OLS estimator of coefficients to be unbiased in multivariate regression?
+ Why is it often the case that the variable of your interest in observational data (i.e., non-experimental data) is correlated with the error term?
+ What does a randomization of a variable do? What is the benefit?


---
class: middle

# Omitted variables and multicollinearity

Conside the model:

$$y_i = \beta_0 + \beta_1 x_{1,i} + \beta_2 x_{2,i} + u_i$$

You are considering the following two models to estimate:

$EE_1$: $y_i=\beta_0 + \beta_1 x_{1,i} + v_i (\beta_2 x_{2,i} + u_i)$

$EE_2$: $y_i=\beta_0 + \beta_1 x_{1,i} + \beta_2 x_{2,i} + u_i$

<br>

.content-box-green[**Questions**]

For each of the fours cases below, answer the two questions below.

+ Is $E[\hat{\beta_1}] = \beta_1$ (unbiasedness) in EE1?
+ In which of EE1 and EE2, $\beta_1$ is more accurately estimated $(Var[\hat{\beta_1}]$ is smaller)?

**Case 1**: $\beta_2=0$, $x_2$ is <span style="color:blue">not</span> correlated with $x_1$?

**Case 2**: $\beta_2=0$ and $x_2$ is <span style="color:blue">highly</span> correlated with $x_1$

**Case 3**: $\beta_2\ne 0$, $x_2$ is <span style="color:blue">not</span> correlated with $x_1$?

**Case 4**: $\beta_2\ne 0$, $x_2$ is <span style="color:blue">highly</span> correlated with $x_1$?

---
class: middle

# Testing

Consider the following model:

$$
\begin{aligned}
income = \alpha + \beta_1 educ + \beta_2 exper + \beta_3 IQ + \beta_4 educ \times female + \beta_5 educ \times IQ + v
\end{aligned}
$$

.content-box-green[**Questions**]

+ What test do you use to test the null hypothesis that `educ` and `exper` have no effect on `income` jointly?
+ What test do you use to test the null hypothesis that the marginal impact of education on income is the same between male and femals workers?
+ Write the t-statics formula for the null hypothesis that the marginal impact of education on income is 0 when IQ is 120. 

---
class: middle

# Homeskedastic error and non-heteroskedastic error

.content-box-green[**Questions**]

+ What happens to your <span style = "color: blue;"> coefficient </span> estiamation using OLS when error is not homoskedastic? Does it introduce bias?
+ What happens to your estimation of the <span style = "color: blue;"> variance of the coefficient estimator </span> if the default variance estimator is used even when the error is not homoskedastic? What would be potential consequences of getting the estimation of the variance of the coefficient estimator wrong?
+ You suspect that your error not homoskedastic, but heteroskedastic. What would you do to deal with this problem?
+ You suspect that your error is clusterd by group. What would you do to deal with this problem?
+ What happens if you do not treat clustered error? 

---
class: middle

# $R^2$

.content-box-green[**Questions**]

+ What does $R^2$ measure?
+ Suppose you are interested in understanding the causal impact of a variable on the dependent variable. Should you care about $R^2$? Does a high value of $R^2$ guarantee that your estimation of the causal impact of the variable is reliable?
+ Suppose you are interested in predicting the dependent variable because you want to use the predicted values for some other purposes. Should you care about $R^2$?



