---
title: "Final Paper Project Expectation"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    self_contained: true
    nature:
      ratio: 4:3
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, center, middle
name: dgp

```{r, child = './../setup.Rmd'}
```

```{r echo = F, eval = F}
setwd("/Users/tmieno2/Dropbox/TeachingUNL/AppliedEconometrics_MS/LectureNotes/Endogeneity/")
```

```{r additional-libraries, include = F, cache = F}
#--- load packages ---#
library(broom)
library(fixest)
library(lmtest)
library(Cairo)
library(tidyverse)
library(haven)
library(extrafont)
library(lubridate)
library(ggrepel)
library(ggpubr)
```

# Final Paper Project Expectation

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---
class: middle

# Introduction (3 points)

+ clear identification of what you are trying to find out (research question) [1.5 points]
+ why the research question is worthwhile answering [1.5 points]

---
class: middle

# Data description and exploration (7 points)

+ the nature of the data with summary statistics table [2 point]
+ visualize a few key variables in a <span style="color:blue"> meaningful </span> way [5 points]

.content-box-green[**Important**]

Do not just present summary statistics and graphs. Discuss what you get out of them. Without any such discussions, you will get $0$.

---
class: middle

# Econometric methods (35 points)

The <span style = "color: blue;"> process and thought history </span> of how you end up with the final econometric models and methods. 

---
class: middle

# Results, Discussions, and Conclusions (5 points)

+ interpret and describe the results [3 points]
+ conclusions [2 points]

---
class: middle

# Details on the Econometric Method Section

.content-box-green[**Data generating process**]
+ What variables are involved in explaining the dependent variable? 
  - List all the variables both observed and <span style = "color: blue;"> unobserved </span>.
+ How are they related with each other? 
  - Multicollinearity?
+ What would be the appropriate functional form? 
  - Non-linear impact (e.g., quadratic, log)? 
  - Interactions terms? 
  - Structural difference?

---
class: middle

# Details on the Econometric Method Section

.content-box-green[**Endogeneity and Econometric Methods**]

+ Extensive discussions on why you may have endogeneity problems. What are the sources? 
  - Omitted variable? 
  - Selection bias? 
  - Reverse causality? 
  - Measurement errors?
+ Hypothetical discussion of what econometric methods you can use to deal with the endogeneity problem 
  - Randomization of the variable of interest? 
  - Panel data approach? 
  - Instrumental variable approach?
  - Difference in difference?
+ Detailed descriptions of what you end up doing (explain the process)
  - Why do you think your approach address the endogeneity problem you identified above or not? (<span style = "color: blue;"> It is not at all a problem that you cannot solve the endogeneity problem entirely </span>)
+ Use the appropriate standard error estimation approach 
(heterogeneity, clustered error, etc)



---
class: middle

# Key to writing a successful paper (high grade)

+ Justify and explain <span style = "color: blue;"> everything </span> you did in the paper!!
  - I tested the joint statistical significance of these interactions terms because ...
+ <span style = "color: blue;"> Re-emphasized: </span> I do not care about your results. What I care is the <span style = "color: blue;"> process </span>!! 
