---
title: "The Impacts of Irrigation Costs on Irrigation Demand"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 4:3
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r, child = './../setup.Rmd'}
```

```{r echo = F, eval = F}
setwd("/Users/tmieno2/Dropbox/TeachingUNL/AppliedEconometrics_MS/LectureNotes/Demonstration/")
```

```{r additional-libraries, include = F, cache = F}
#--- load packages ---#
suppressMessages(library(broom))
suppressMessages(library(fixest))
suppressMessages(library(flextable))
suppressMessages(library(officer))
suppressMessages(library(gt))
suppressMessages(library(lmtest))
suppressMessages(library(lfe))
suppressMessages(library(tidyverse))
suppressMessages(library(data.table))
```

# Econometric Model (univariate)

$W = \beta_0 + \beta_1 IC + v$

+ $W$: irrigation amount 
+ $IC$: irrigation cost 
+ $v$: error term.

---
class: middle

# Endogeneity: Omitted Variable Bias

$W = \beta_0 + \beta_1 IC + v$

.content-box-green[**Question**]

What is in $v$? Are they going to cause bias on $\beta_1$ estimation? 

--

+ soil type
+ well yield
+ precipitation 

---
class: middle

# Econometric Model (multi-variate) 

$W = \beta_0 + \beta_1 IC+ \beta_2 WY + \sigma_5 P^5 + \sigma_6 P^6 + \sigma_7 P^7 + \sigma_8 P^8 + \sigma_9 P^9 + \varepsilon$

+ $W$: irrigation amount 
+ $IC$: irrigation cost 
+ $WY$: well yield
+ $P^x$: total precipitation in month $x$
+ $\varepsilon$: error term

---
class: middle

# Panel Data Method

$W_{i,t} = \beta_0 + \beta_1 IC_{i,t}+ \beta_2 WY_{i,t} + \sigma_5 P^5_{i,t} + \sigma_6 P^6_{i,t} + \sigma_7 P^7_{i,t} + \sigma_8 P^8_{i,t} + \sigma_9 P^9_{i,t} + \varepsilon_{i,t}$

+ $W$: irrigation amount 
+ $IC$: irrigation cost 
+ $WY$: well yield
+ $P^x$: total precipitation in month $x$
+ $\varepsilon$: error term

---
class: middle

# Endogeneity

.content-box-red[**Declining block rate pricing**]

```{r echo = F}
fread("REA_price_schedule.csv") %>%
  .[REA == "Highline" & year >= 2013, t2 := NA] %>%
  .[REA == "Highline" & year >= 2013, p3 := NA] %>%
  .[year <= 2016, ] %>%
  .[, REAnum := NULL] %>%
  .[, REA := NULL] %>%
  setnames(names(.), c("Year", "Price 1", "Price 2", "Price 3", "Threshold 1", "Threshold 2")) %>%
  .[, Supplier := "Highline"] %>%
  .[7:12, Supplier := "Y-W"] %>%
  relocate(Supplier) %>%
  flextable() %>%
  merge_v(j = 1) %>%
  hline(
    i = 6,
    border = fp_border(width = 1)
  ) %>%
  fix_border_issues() %>%
  autofit()
```

--

.content-box-green[**Example**]

A well in Highline’s service area with 100 HP will be on the second price tier after using 40,000 kWh.

---
class: middle

# Endogeneity

.content-box-red[**Declining block rate pricing**]

```{r echo = F}
fread("REA_price_schedule.csv") %>%
  .[REA == "Highline" & year >= 2013, t2 := NA] %>%
  .[REA == "Highline" & year >= 2013, p3 := NA] %>%
  .[year <= 2016, ] %>%
  .[, REAnum := NULL] %>%
  .[, REA := NULL] %>%
  setnames(names(.), c("Year", "Price 1", "Price 2", "Price 3", "Threshold 1", "Threshold 2")) %>%
  .[, Supplier := "Highline"] %>%
  .[7:12, Supplier := "Y-W"] %>%
  relocate(Supplier) %>%
  flextable() %>%
  merge_v(j = 1) %>%
  hline(
    i = 6,
    border = fp_border(width = 1)
  ) %>%
  fix_border_issues() %>%
  autofit()
```

--

.content-box-green[**Endogeneity?**]

Would this cause bias in estimating $\beta_1$?

---
class: middle

# Fixed-Effects estimation 

```{r echo = F}
massage_data_2p <- data.frame(
  Location = rep(c("Chicago", "Peoria", "Milwaukee", "Madison"), each = 2),
  Year = rep(c(2003, 2004), 4),
  P = c(75, 85, 50, 48, 60, 65, 55, 60),
  Q = c(2, 1.8, 1.0, 1.1, 1.5, 1.4, 0.8, 0.7),
  Ql = rep(c(10, 5, 7, 6), each = 2)
)
```

```{r echo = F}
massage_data_2p %>%
  flextable()
```

---
class: middle

# (individual) Fixed Effects Estimation

+ Including individual (cross-sectional units) dummies (here, it is well represented by `wdid`) in the model

+ This is equivalent to do within-transformation of the data (deviation from the individual means) and then regress deviation on deviation

+ What the within-transformation does is to effectively  divide the observations into groups (wells) and you look at variations .red[within] each of the groups (wells observed over time).

+ Since including individual dummies are equivalent to the within-transformation approach, that means including individual (well) dummies also effectively divides the observations into groups (wells) and you look at variations .red[within] each of the groups (wells observed over time).

---
class: middle

# Fixed Effect Estimation (in general)

.content-box-green[**key**]

Including dummies variables effectively divide the observations into groups and you look at variations .red[within] each of the groups

---
class: middle

# Beyond individual (well) FEs  

```{r echo = F}
final_data <- readRDS("final_data_CO.rds")

final_data[wdid %in% c(4905049, 4905044), .(wdid, year, pumpingAF, pc, tier, pump_hp)] %>%
  arrange(wdid, year) %>%
  flextable() %>%
  merge_v(j = 1) %>%
  hline(
    i = 5,
    border = fp_border(width = 1)
  ) %>%
  fix_border_issues() %>%
  autofit()
```

--

.content-box-green[**Question**]

+ Does including individual dummies (well fixed effects) help?

--

+ Does including well-tier dummies (well-tier fixed effects) help? 

--

+ Does including well-tier-hp dummies (well-tier-hp fixed effects) help? 
